Data exploration, quality control and statistical analysis of ChIP-exo experiments

### 2016-05-03 rmarkdown/paper.Rnw
added enrichment and library complexity section to paper, replaced respective figure with more comprehensive analysis

% It is calculated as the correlation between both strand
% coverages, where each one is shifted $\delta / 2$ bp towards the
% $3\prime$ direction
\color{blue}

@@

\begin{enumerate}
\item \st{Available data, the motivation of this part is two fold:}
 \begin{itemize}
 \item \st{we want to list all data sets we obtained}
 \item \st{we want to emphasize that for the sig70 datasets we have a
   unique design that allows us to compare chip exo with SE and PE
   chipseq , the later one for the first time}
 \end{itemize}
\item \st{comparison with chip seq} %? with and without dpeak ?}
\item \st{only in methods. list the align algorithm used}
\item \st{chip seq qc metrics}
\item chip exo qc pipeline, description and justification of the steps
\begin{itemize}
\item need to justify the sampling step. for this we may use the depth
  table and add the factor to normalize one depth to another. explain
  the confounding of pcr amplification and the reduction of the number
  of possible mappable positions.
\item case study with FoxA1, analysis of enrichment vs library
  complexity and analysis and the strand imbalance tools. in methods
  we need to add which tests did we perform to show that the pipeline
  works. how do we expect from the diagnostics and how to interpret
  the plots.
\item analysis of why the pipeline works. in methods need to specify
  how and in results need to specify why it works, and what are the
  conclusions from it
\end{itemize}
\item high resolution analysis with chip exo and comparison with PE
  chip seq
  \begin{itemize}
  \item justify what do we used dpeak (in methods need to list the
    other methods used). perhaps we can add gof plots in
    supplement. we also want to compare with the same method,
    therefore we can use the fact that dpeak have a unified model for
    PE and SE dpeak
  \item after comparing with other methods we can compare with PE and
    SE ChIP-seq, this comparison gives also resolution and sensitivity
  \item compare samples under same budget of reads
  \end{itemize} 
\end{enumerate}

@@

\subsubsection{ChIP-exo Quality Control Pipeline}

Figure \ref{fig:qcdiagram} shows a flowchart for the ChIP-exo QC
pipeline. In the first step, we partition the genome by keeping the
non-digested ChIP-exo regions. Then, for each region it counts the
number of fragments that compose the region and the number of
positions to which the reads are being mapped to in each strand. With
these values it calculates the following summary statistics:

Finally it creates several visualizations designed to diagnose the
quality level of a ChIP-exo sample. Figure \ref{fig:qcdiagram}A shows
the typical behavior of the ARC vs. URCR plot. In general, the plot
depicts two strong arms: One on the left with low ARC values and
varying URCR which corresponds to ChIP-exo's background, regions that
are usually composed by scattered reads that were not digested during
the exonuclease step; and another one where the URCR decreases as the
ARC increases, which corresponds to regions that are usually enriched
and as the URCR decreases the library complexity does it as well, on
the other hand high URCR values correspond to regions composed by
position with relatively few fragments aligned to. Figures
\ref{fig:qcdiagram}B and \ref{fig:qcdiagram}C analyze the strand
imbalance bias in a ChIP-exo experiment: The first one depicts how
quickly the regions exclusively formed by fragments in one strand are
being filtered out as regions with higher depth are observed; and the
second one shows how quickly the FSR's distribution approach the
median, since in a high quality sample it is expected for the median
to be approximately 0.5 and the enriched regions are going to be
composed by fragments sequenced from both strands.

The ChIP-exo QC pipeline provides a model free framework to analyze
the biases in a ChIP-exo experiment by taking advantage of the
exonuclease enzyme that digests the non-enriched regions to partition
the genome, calculates common ChIP-Seq QC metrics in ChIP-exo regions
locally and allows the interpretation of these metrics by the use of
diagnostic plots.

\textbf{Enrichment and library complexity in ChIP-exo data}

Using the Fox A1 in mouse liver cell lines from \cite{exoillumina} and
these two quantities, we explored the relationship between library
complexity and experiment enrichment. In Figure \ref{fig:enrich}A we
present ARC vs. URCR plots for all three replicates. As a case of
study, we compare the three plots to differentiate the quality of the
three experiment. Hence, this might imply that the first replicate to
have have more enriched regions and the third replicate's library
complexity to be lower than the other two replicates library
complexities. To verify this statement, we extracted the sequences
around high confidence binding events and look for the FoxA1 motif
using FIMO \cite{fimo}. Figure \ref{fig:enrich}B shows the number of
candidate regions, which shows that the first replicate is being
allocated into more enriched regions than the other ones. Figure
\ref{fig:enrich}C shows that for the first and third replicates, the
FoxA1 motif is being detected in roughly the same proportion of sites,
and finally in \ref{fig:enrich}D we observe that the first and third
replicate can detect the FoxA1 motif with the same significance, while
the second replicate does not.

\textbf{Strand imbalance in ChIP-exo data}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width = .9  \textwidth,page = 1]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
  \newline
  \includegraphics[width = .3 \textwidth,page = 1]{../figs/FOXA1_mm9_fimo/FOXA1_summaries_by_replicate.pdf}
  \includegraphics[width = .3 \textwidth,page = 3]{../figs/FOXA1_mm9_fimo/FOXA1_summaries_by_replicate.pdf}
  \includegraphics[width = .3 \textwidth,page = 1]{../figs/FOXA1_mm9_fimo/FOXA1_log10pval_cumulative_nmotifs.pdf}
  \caption{Using the mouse FoxA1 experiment from \cite{exoillumina}:
    A) Hexbin plots of $\mbox{ARC}$ against $\mbox{URCR}$, there is a
    slight separation into two strong arms, one corresponds to low
    $\mbox{ARC}$ and varying $\mbox{URCR}$, and for the other
    $\mbox{URCR}$ decreases as $\mbox{ARC}$ increases. B) Number of
    candidate sites for each replicate. C) Percentage of candidate
    sites where the FoxA1 motif was detected. D) Cumulative
    proportion of detected motifs by replicate.}
  \label{fig:enrich}
\end{figure}

>>>>

% It is calculated as the correlation between both strand
% coverages, where each one is shifted $\delta / 2$ bp towards the
% $3\prime$ direction

\subsubsection*{ChIP-exo Quality Control Pipeline}

We first show the overall pipeline and consecutively discuss
individual components with a case study using the FoxA1 factor in
mouse liver cell lines generated by Serandour et al., 2013, and
finally we show the evaluation of the ChIP-exo and ChIP-nexus samples
with the QC pipeline. Figure \ref{fig:qcdiagram} shows a flowchart for
the ChIP-exo QC pipeline. Given a sample of ChIP-exo or ChIP-nexus
reads, in the first step, we partition the genome by keeping the
non-digested ChIP-exo regions. Then, for each region it counts the
number of fragments that compose the region and the number of
positions to which the reads are being mapped to in each strand. With
these values it calculates the following summary statistics:

Then it creates several visualizations designed to diagnose the
quality level of a ChIP-exo sample. Figure \ref{fig:qcdiagram}A shows
the typical behavior of the ARC vs. URCR plot. In general, the plot
depicts two strong arms: One on the left with low ARC values and
varying URCR which corresponds to ChIP-exo's background, regions that
are usually composed by scattered reads that were not digested during
the exonuclease step; and another one where the URCR decreases as the
ARC increases, which corresponds to regions that are usually enriched
and as the URCR decreases the library complexity does it as well, on
the other hand high URCR values correspond to regions composed reads
aligned to few positions. Finally we quantify we quantify the
relationship between library complexity by the use of two indexes that
represent the change in the number of unique positions per regions and
the change in of the width of a regions as the depth changes. Figures
\ref{fig:qcdiagram}B and \ref{fig:qcdiagram}C analyze the strand
imbalance bias in a ChIP-exo experiment: The first one depicts how
quickly the regions exclusively formed by fragments in one strand are
being filtered out as regions with higher depth are observed, in a
high quality sample the proportion of regions composed by only one
strand is expected to decrease as higher depth regions are considered,
while in a lower quality dataset this proportions remains
approximately constant; and the second one shows how quickly the FSR's
distribution approach the median, since in a high quality sample it is
expected for the median to be approximately 0.5 and the enriched
regions are going to be composed by fragments sequenced from both
strands.

% The ChIP-exo QC pipeline provides a model free framework to analyze
% the biases in a ChIP-exo experiment by taking advantage of the
% exonuclease enzyme that digests the non-enriched regions to partition
% the genome, calculates common ChIP-Seq QC metrics in ChIP-exo regions
% locally and allows the interpretation of these metrics by the use of
% diagnostic plots.

\textbf{Analysis of enrichment and library complexity in FoxA1
  ChIP-exo data}

Using the Fox A1 in mouse liver cell lines generated by Serandour et
al., 2013\nocite{exoillumina} and these two quantities, we explored
the relationship between library complexity and experiment
enrichment. In Figure \ref{fig:enrich}A we present ARC vs. URCR plots
for all three replicates. The first and third replicate show a defined
decreasing trend in the $\mbox{URCR}$ as the $\mbox{ARC}$ increases
which indicates that this samples exhibit a higher enrichment than the
second replicate. On the other hand, the overall $\mbox{URCR}$ level
from the first two replicates is higher than the third replicate's
level, which indicates that its libraries are more complex than the
third replicate since three experiment are comparable in depth.

Using the regions and summary statistics we obtained a set of
candidate sites for each replicate, the total number of candidate
sites is shown in Figure \ref{fig:enrich}B. For this sites, we
extracted the sequence and searched for the FoxA1 motif using
\textbf{FIMO} \cite{fimo}. Figure \ref{fig:enrich}C shows the fraction
of candidate site where a motif was found. The first replicate
outperforms the rest in both number of candidate sites and proportion
of sites with motif; on the other hand we can see that the second
replicate candidate sites are more likely to contain the motif
sequence, but there more candidate sites in the third replicate than
in the second. Finally, Figure \ref{fig:enrich}D shows the average
profiles around the coordinate where the motif was found in the
candidate sites: We can observe that the signal is more delimited for
the first and third replicates than the second one, but overall the
signal pattern is reproducible in the three ChIP-exo
experiments. Figures \ref{fig:enrich}E, \ref{fig:enrich}F and
\ref{fig:enrich}G show the purity of the motif found by visualizing
the sequence to which the motif was matched. All these observations
together shows the following findings: First, observing a defined
decreasing trend in $\mbox{URCR}$ as $\mbox{ARC}$ increases implies
than the sample is more enriched, therefore a higher amount of
candidate sites. Second, for experiments that are not deeply
sequenced, low $\mbox{URCR}$ values represent low complexity which
decreases the amount of sites of motif found but it doesn't modify the
read distribution around the motif. Finally, libraries with higher
complexity returns sequences with a better motif match.

% To verify this statement, we
% extracted the sequences around high confidence binding events and look
% for the FoxA1 motif using FIMO \cite{fimo}. Figure \ref{fig:enrich}B
% shows the number of candidate regions, which shows that the first
% replicate is being allocated into more enriched regions than the other
% ones. Figure \ref{fig:enrich}C shows that for the first and third
% replicates, the FoxA1 motif is being detected in roughly the same
% proportion of sites, and finally in \ref{fig:enrich}D we observe that
% the first and third replicate can detect the FoxA1 motif with the same
% significance, while the second replicate does not.
\color{blue}

@@

\begin{enumerate}
\item \st{Available data, the motivation of this part is two fold:}
 \begin{itemize}
 \item \st{we want to list all data sets we obtained}
 \item \st{we want to emphasize that for the sig70 datasets we have a
   unique design that allows us to compare chip exo with SE and PE
   chipseq , the later one for the first time}
 \end{itemize}
\item \st{comparison with chip seq} %? with and without dpeak ?}
\item \st{only in methods. list the align algorithm used}
\item \st{chip seq qc metrics}
\item \st{chip exo qc pipeline, description and justification of the steps}
\begin{itemize}
\item need to justify the sampling step. for this we may use the depth
  table and add the factor to normalize one depth to another. explain
  the confounding of pcr amplification and the reduction of the number
  of possible mappable positions.
\item \st{case study with FoxA1, analysis of enrichment vs library
  complexity and analysis and the strand imbalance tools. in methods
  we need to add which tests did we perform to show that the pipeline
  works. how do we expect from the diagnostics and how to interpret
  the plots.}
\item analysis of why the pipeline works. in methods need to specify
  how and in results need to specify why it works, and what are the
  conclusions from it
\end{itemize}
\item high resolution analysis with chip exo and comparison with PE
  chip seq
  \begin{itemize}
  \item justify what do we used dpeak (in methods need to list the
    other methods used). perhaps we can add gof plots in
    supplement. we also want to compare with the same method,
    therefore we can use the fact that dpeak have a unified model for
    PE and SE dpeak
  \item after comparing with other methods we can compare with PE and
    SE ChIP-seq, this comparison gives also resolution and sensitivity
  \item compare samples under same budget of reads
  \end{itemize} 
\end{enumerate}

@@

\textbf{Analysis of strand imbalance in FoxA1 ChIP-exo data}

@@

\begin{figure}[h!]
  \centering
  \includegraphics[width = .9  \textwidth,page = 1]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
  \newline
  \includegraphics[width = .23 \textwidth, page =
  1]{../figs/for_paper/FoxA1_barplots.pdf}
  \includegraphics[width = .23 \textwidth, page =
  3]{../figs/for_paper/FoxA1_barplots.pdf}
  \includegraphics[width =
  .4\textwidth]{../figs/for_paper/FoxA1_profiles_around_motif.pdf}
  \includegraphics[width = .3\textwidth,page =
  2]{../figs/for_paper/FoxA1_matched_motif_sequence.pdf}
  \includegraphics[width = .3\textwidth,page =
  3]{../figs/for_paper/FoxA1_matched_motif_sequence.pdf}
  \includegraphics[width = .3\textwidth,page =
  1]{../figs/for_paper/FoxA1_matched_motif_sequence.pdf}
  % \includegraphics[width = .3 \textwidth,page = 1]{../figs/FOXA1_mm9_fimo/FOXA1_summaries_by_replicate.pdf}
  % \includegraphics[width = .3 \textwidth,page = 3]{../figs/FOXA1_mm9_fimo/FOXA1_summaries_by_replicate.pdf}
  % \includegraphics[width = .3 \textwidth,page = 1]{../figs/FOXA1_mm9_fimo/FOXA1_log10pval_cumulative_nmotifs.pdf}
  % \caption{Using the mouse FoxA1 experiment from \cite{exoillumina}:
  %   A) Hexbin plots of $\mbox{ARC}$ against $\mbox{URCR}$, there is a
  %   slight separation into two strong arms, one corresponds to low
  %   $\mbox{ARC}$ and varying $\mbox{URCR}$, and for the other
  %   $\mbox{URCR}$ decreases as $\mbox{ARC}$ increases. B) Number of
  %   candidate sites for each replicate. C) Percentage of candidate
  %   sites where the FoxA1 motif was detected. D) Cumulative
  %   proportion of detected motifs by replicate.}
  \caption{Using the mouse FoxA1 experiment from \cite{exoillumina}:
    A) Hexbin plots of $\mbox{ARC}$ against $\mbox{URCR}$, there is a
    slight separation into two strong arms, one corresponds to low
    $\mbox{ARC}$ and varying $\mbox{URCR}$, and for the other
    $\mbox{URCR}$ decreases as $\mbox{ARC}$ increases. B) Number of
    candidate sites for each replicate. C) Percentage of candidate
    sites where the FoxA1 motif was detected. D) Average coverage
    around FoxA1 motif. Base distribution for matched sequence for
    Rep-1 (E), Rep-2 (F) and Rep-3 (G).}
  \label{fig:enrich}
\end{figure}

{\color{red} \emph{need to add labels for each figure and think about
    order, in the supplement we have ecdf from fimo score or pval, and
    overlaps with peaks called by mosaics}}
